---
title: "R Programming Week 1, Subsetting Data and Vectorized Operations"
author: "Krista DeStasio"
date: "12/26/2016"
output: html_document
---
# Notes from R Programming, week 1
*Note: much of the following text is directly from the course slides, available here: [www.coursera.org/learn/r-programming](www.coursera.org/learn/r-programming)*
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("~/Dropbox (PfeiBer Lab)/kdestasio/R_projects/coursera/r_programming/")
rm(list = ls())
list.of.packages <- c("datasets")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
library(datasets)
```
# SUBSETTING IN R
______________________________
## Subsetting Basics
There are a number of operators that can be used to extract subsets of R objects.

- [ always returns an object of the same class as the original; can be used to select more than one element
- [[ is used to extract elements of a list or a data frame; it can only be used to extract a single element and the class of the returned object may not necessarily be a list or a data frame
- $ is used to extract elements of a list or a data frame by name; semantics are similar to that of [[
```{r}
x <- c("a", "b", "c", "c", "d", "a")
```
Examples subsetting using a NUMERIC index
```{r}
x[1]
x[2]
x[1:4]
```
```{r}
# Examples subsetting using a LOGICAL index
x[x > "a"]
(u <- x > "a")
x[u]
```
## Subsetting Lists
```{r}
x <- list(foo = 1:4, bar = 0.6)
x[1] # Returns a list because [ always returns the element of the same class as the original
x[[1]] # [[ Returns just the sequence
x$bar # Extracts the element "bar" from the list
x[["bar"]] # The same as x$bar
x["bar"] # Returns a list wit the element bar in it
```
To extract multiple elements of a list, you need to use the single bracket operator.
```{r}
x <- list(foo = 1:4, bar = 0.6, bax = "hello")
x[c(1, 3)]
```
The [[ operator can be used with *computed* indices; $ can only be used with literal names.
```{r}
name <- "foo"
x[[name]] # computed index for "foo" because that's the value of the "name" variable
x$name # $ literally looks for an element of the list that has the name "name" associated with it
x$foo
```
### Subsetting Nested Elements of a List
The [[ can take an integer sequence.
```{r}
(x <- list(a = list(10, 12, 14), b = c(3.14, 2.81)))
x[[c(1, 3)]]
x[[1]][[3]]
x[[c(2, 1)]]
```
## Subsetting a Matrix
Matrices can be subset in the usual way
```{r}
(x <- matrix(1:6, 2, 3))
x[1, 2]
x[2, 1]
```
Indices can also be missing.
```{r}
(x <- matrix(1:6, 2, 3))
x[1, ]
x[, 2]
```
By default, when a single element of a matrix is retrieved, it is returned as a vector of length 1 rather than a 1x1 matrix. This behavior can be turned off by setting *drop* = FALSE.  

When I subset one element of a matrix with [ I don't get back a matrix unless I use *drop*
```{r}
(x <- matrix(1:6, 2, 3))
x[1, 2]
x[1, 2, drop = FALSE] # drop = FALSE does not drop the dimensions
```
Subsetting a single column or a single row will give you a vector, not a matrix (by default).
```{r}
(x <- matrix(1:6, 2, 3))
x[1, ]
x[1, , drop = FALSE]
```
## Subsetting, Partial Matching
Partial matching of names is allowed with [[ and $
```{r}
x <- list(aardvark = 1:5)
x$a # Since there is only one element that starts with "a", we get the object associated with "aardvark" back
x[["a"]] # [[ expects to get an exact match back
x[["a", exact = FALSE]] # allows for partial matching
```
## Removing Missing Values
Most data have missing values. For a vector, matrix, or data frame, you want to create a logical vector that tells you where the missing data (NAs) are.
```{r}
x <- c(1, 2, NA, 4, NA, 5)
bad <- is.na(x) # Creates a logical vector to locate the missing data
x[!bad] # Will return the opposite of "bad", giving me the good elements
```
What if there are multiple things and you want to take the subset with no missing values?
```{r}
x <- c(1, 2, NA, 4, NA, 5)
y <- c("a", "b", NA, "d", NA, "f")
good <- complete.cases(x, y)
good
x[good]
y[good]
```
### Removing Missing Values From Data Frames
```{r}
airquality[1:6. ]
good <- complete.cases(airquality)
airquality[good, ][1:6, ]
```
# VECTORIZED OPERATIONS
__________________________
The idea with vectorized operations is that things can happen in parallel. Many operations in R are *vectorized* making code more efficient, concise, and easier to read.
```{r}
x <- 1:4; y <- 6:9
x + y
x > 2
x >= 2
y ==8
x * y
x / y
```
Can also do vectorized matrix operations.
```{r}
(x <- matrix(1:4, 2, 2))
(y <- matrix(rep(10, 4), 2, 2))
# Element-wise computations
x * y
x / y
# True matrix computation
x %*% y
```