---
title: "R Programming Week 2, Control Structures"
author: "Krista DeStasio"
date: "12/26/2016"
output: html_document
---
# Notes from R Programming, week 1
*Note: much of the following text is directly from the course slides, available here: [www.coursera.org/learn/r-programming](www.coursera.org/learn/r-programming)*
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("~/Dropbox (PfeiBer Lab)/kdestasio/R_projects/coursera/r_programming/")
rm(list = ls())
list.of.packages <- c("datasets")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
library(datasets)
```
# CONTROL STRUCTURES
__________________________
## Introduction
Control structures in R allow you to control the flow of execution of the program, depending on the runtime conditions. Common structures are:

- **if, else**: testing a condition
- **for**: execute a loop a fixed number of times
- **while**: execute a loop a fixed number of times
- **repeat**: execute an infinite loop
- **break**: break the exeecution of a loop
- **next**: skip an iteration of a loop
- **return**: exit a function

Most control structures are not used in interactive sessions, but rather when writing functions or longer expressions.

## if-else
This structure allows for dependencies in a function.  
There can be any number of *if-else* constructs.  
There are a couple different ways to construct the *if-else* construct.  
```{r, eval=FALSE}
# This is a valid if-else construct
if(x > 3) {
        y <- 10
} else {
        y <- 0
}
# And so is this
y <- if(x > 3) {
        10
} else {
        0
}
```
The *else* clause is not necessary. Can use *if* clause by itself.

## For Loops
*for* loops take an interator variable and assign it successive values from a swquence or vector. For loops are most commonly used for iterating over the elements of an object (list, vector, etc.)
```{r, eval=FALSE}
for(i in 1:10) {
  print(i)
}
```
This loop takes the **i** variable in each iteration of the loop and gives it values 1, 2, 3, ..., 10, and then exits.  

These three loops have the same behavior.
```{r}
x <- c("a", "b", "c", "d")
# All of the following for loops are equivalent
# The first example is in a format that is common in other languages
for(i in 1:4){
        print(x[i])
}
# The seq_along function takes a vector as the input and creates an integer sequence that is equal to the length of that vector
for(i in seq_along(x)) {  # Generates a vector the length of variable x
        print(x[i])
}
# "letter" is the index variable and takes values from the vector itself
for(letter in x) {
        print(letter)
}
# The same as the first example, but without the curly braces
for(i in 1:4) print(x[i])
```
## Nested for Loops
```{r}
x <- matrix(1:6, 2, 3)
# For 2-dimensional matrix, can loop over rows and then columns
for(i in seq_len(nrow(x))) {    # Loop over the rows of length x and create an integer sequence
      for(j in seq_len(ncol(x))) {    # Use the number of columns to create an integer sequence
              print(x[i, j])
      }
}
```
Be careful with nesting. Nesting beyond 2-3 levels is often difficult to read/understand. Can often get around this by using functions.

## while
*while* loops begin by testing an expression. If it is true, then they execute the loop body. Once the loop body is executed, the condition is tested again, and so on.
```{r, eval=FALSE}
count <- 0
while(count < 10) {
      print(count)
  count <- count + 1
}
```
*while* loops can potentially result in infinite loops, so be careful! Often safer to use something like a *for* loop that has a hard limit on the number of times it will execute.  

Sometimes there will be more than one condition in the test.
```{r, eval=FALSE}
z <- 5
while(z >= 3 && z <= 10) {
      print(z)
  coin <- rbinom(1, 1, 0.5)
  
  if(coin == 1) { ## randome walk
        z <- z + 1
  } else {
          z <- z - 1
        }
}
```
Conditions are always evaluated from left to right.

##Repeat, Next, Break
### repeat, break
*repeat* initiates an infinite loop; these are not commonly used in statistical applications, but they do have their uses. The only way to exit a *repeat* loop is to call *break*.  

It's better to set a hard limit on the number of iterations (e.g. for using a loop) and then report whether convergence was achieved or not.  

### next, return
*next* is used to skip an iteration of a loop
```{r eval=FALSE}
for(i in 1:100) {
  if(i <= 20) {
    # Skip the first 20 iterations
    next
  }
  # Do something here
}
```
*return* signals that a function should exit and return a given value.  