---
title: "R Programming Week 2, Coding Standards"
author: "Krista DeStasio"
date: "12/29/2016"
output: html_document
---
# Notes from R Programming, week 1
*Note: much of the following text is directly from the course slides, available here: [www.coursera.org/learn/r-programming](www.coursera.org/learn/r-programming)*
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("~/Dropbox (PfeiBer Lab)/kdestasio/R_projects/coursera/r_programming/")
rm(list = ls())
list.of.packages <- c("datasets", "ggplot2")
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
library(datasets)
library(ggplot2)
```
# CODING STANDARDS
___________________

1. Always use text files / text editor
    + Can be read by pretty much any program
    + The editor in Rstudio is a text editor
2. Indent your code
3. Limit the width of your code (80 columns?)
    + Improves readability
    + Fixing line length (80 columns) prevents lots of nesting and very long functions
   + Suggested: indents of 4 spaces at minimum
Suggested 
4. Limit the length of individual functions
    + A function should do one thing
    + Helpful for readability and debugging
    
# DATES AND TIMES IN R
_______________________
R has a special way o represent dates and times

- Dates are represented by the **Date** class
- Times are represented by the **POSIXct** or the **POSIXlt** class
- Dates are stored internally as the number of days since 1970-01-01
- Times are stored internally as the number of seconds since 1970-01-01

## Dates in R
Dates are represented by the *Date* class and can be coerced from a character string using the **as.Date( )** function
```{r}
x <- as.Date("1970-01-01")
x # Will print out a character string
unclass(x) # Unclass the object to get a number
unclass(as.Date("1970-01-02"))
```
Dates are stored as *objects* of the *Date class*.

## Times in R
POSIX is a family of computing standards for how to represent dates and times

- in the POSIXct class, times are stored as a very large integer under the hood; it is a useful class when you want to sore something like a data frame
- POSIXlt is a list underneath and it stores a bunch of other useful information like the day of the week, day of the year, month, day of the month

There are a number of generic functions that work on dates and times

- **weekdays**: give the day of the week
- **months**: give the month name
- **quarters**: give the quarter number (Q1, Q2, Q3, Q4)

Times can be coerced from a character string using the **as.POSIXlt** or **as.Posixct** functions
```{r}
x <- Sys.time()
x
p <- as.POSIXlt(x)
names(unclass(p))
p$sec
```
There is also the **strptime** function in case your dates are written in a different format
```{r}
datestring <- c("January 10, 2012 10:40", "December 9, 2011 9:10")
x <- strptime(datestring, "%B %d, %Y %H:%M ")
x
class(x)
```
To check the formatting strinfs: **?strptimes** for details

## Operations on Dates and Times

You can use + and - on dates and times. You can also do comparisons (i.e. ==, <=)
```{r}
x <- as.Date("2012-01-01")
y <- strptime("9 Jan 2011 11:34:21", "%d %b %Y %H:%M:%S")
x-y
x <- as.POSIXlt(x) # Both need to be the same type of object
x-y
```
Even keeps track of leap years, leap seconds, daylight savings, and time zones.
