---
title: "R Programming Week 1, Data Types"
author: "Krista DeStasio"
date: "12/23/2016"
output: html_document
---
# Notes from R Programming, week 1
*Note: much of the following text is directly from the course slides, available here: [www.coursera.org/learn/r-programming](www.coursera.org/learn/r-programming)*
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("~/Dropbox (PfeiBer Lab)/kdestasio/R_projects/coursera/r_programming/")
rm(list = ls())
list.of.packages <- c()
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages)
library()
```
# DATA TYPES
______________________
## R Objects and Attributes:
### Numbers
Numbers in R treated as numeric. If you want an integer, you must specify the L suffix.  

### Attributes
R objects can have attributes:

- names, dimnames
- dimensions (e.g. matrices, arrays)
- class
- length
- metadata

Attributes can be accessed using the **attributes( )** function.

## Vectors and Lists:
### Creating vectors
The **C( )** function:
```{r vectors}
x <- c(1, 2) # numeric
x <- c(TRUE, FALSE) # logical
x <- c(T, F) # logical
x <- c("a", "b", "c") # character
x <- 9:23 # integer
x <- c(1+0i, 2+4i) # complex
```
The **vector( )** function:
```{r more vectors}
(x <- vector("numeric", length = 10))
```
Mixing objects:
```{r}
y <- c(1.2, "a") # character
y <- c(TRUE, 2) # numeric
y <- c("a", TRUE) # characer
```
When different objects are mixed in a vector, *coercion* occurs so that every element in the vector is of the same class.  

### Explicit coercion
Use the **as.** functions, if available.
```{r coercion}
x <- 0:6
class(x)
as.numeric(x)
as.logical(x)
as.character(x)
as.complex(x)
```
Nonsensical coercion will result in NAs.  

### Lists
Type of vector that can contain elements of different classes.
```{r lists}
(x <- list(1, "a", TRUE, 1 + 4i))
```
Elements of a list have double brackets around them while elements of other vectors have single brackets.  

## Matrices:
Matrices are a special type of vector with a *dimension* attribute. The dimension attribute is itself an integer vector of length 2 (nrow, ncol).  
Matrices are constructed column-wise.
```{r matrices}
(m <- matrix(1:6, nrow = 2, ncol = 3))
dim(m)
attributes(m)
```
Matrices can be created directly from vectors by adding a dimension attribute.
```{r more matrices}
(m <- 1:10)
dim(m) <- c(2, 5) # assigning the matrix dimensions: 2 rows, 5 columns
m
```
### cbind and rbind
Matrices can be created by *column-binding* or *row-binding* with **cbind( )** and **rbind( )**
```{r binding}
x <- 1:3
y <- 10:12
cbind(x, y)
rbind(x, y)
```

## Factors:
Represent categorical data (ordered or unordered). Can be thought of as integer vectors where each integer has a label.  

- Treated specially by modeling functions like **lm( )** and **glm( )**
```{r factors}
x <- factor(c("yes", "yes", "no", "no")) # use the factor function to create a facor
table(x)
unclass(x) # strips out the class from the vector
attr(x, "levels")
```
The order of the levels can be set using the **levels** argument to **factor( )**. This can be important in linear modeling because the first level is used as the baseline.  

## Missing Values:
Denoted as NA or NaN. NaN is for undefined mathematical operations.

- **is.na( )** is used to test if objects are NA
- **is.nan( )** is used to test for NaN
- NA values have a class
- A NaN is also NA, but the converse is not true
```{r}
x <- c(1, 2, NA, 10, 3)
is.na(x)
is.nan(x)
x <- c(1, 2, NA, NaN, 4)
is.na(x)
is.nan(x)
```
## Data Frames:
Used to store tabular data

- Represented as a special type of list where every element of the list has the same length
- Each element of the list can be thought of as a column and he length of each element in the list is the number of rows.
- Unlike matrices, data frames can store different types of objects in each column (just like lists); matrices must have every element of the same class
- Data frames also have a special attribute called **row.names**
- Data frames are usually created by calling **read.table( )** or **read.csv( )**
- Can be converted to a matrix by calling **data.matrix( )**
```{r data frames}
(x <- data.frame(foo = 1:4, bar = c(T, T, F, F)))
nrow(x)
ncol(x)
```
## Names Attribute
R objects can also have names.
```{r}
x <- 1:3
names(x)
names(x) <- c("foo", "bar", "narf")
names(x)
```
Lists an also have names.
```{r}
x <- list(a = 1, b = 2, c = 3)
x
```
And matrices can have names.
```{r}
m <- matrix(1:4, nrow = 2, ncol = 2)
dimnames(m) <- list(c("a", "b"), c("c", "d")) # first element of list is vector of row names, second element is vector of column names
m
```
## Summary
Data types

- atomic classes: numeric, logical, character, integer, complex
- vectors, lists
- factors
- missing values
- data frames
- names